### 2016-06-29 doc/publications/plos16/plos16.tex
added reference, some small changes

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
Using VariantSpark, we successfully built a k-means model on the full 2500x80Million matrix to cluster individuals by
their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests (RF) as an ensemble of weak learners 
\marginpar{weak learners only in a certain sense. Originally ``weak learners'' was applied to decision stumps}
overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
between features, which is of importance for modelling complex polygenic diseases.

@@

We therefore we developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
excessive resource requirements on the driver or worker nodes.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the first section we

@@

As discussed in the Introduction, random forests are robust against overfitting and hence can generalise for datasets
where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.

>>>>

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework.
Using VariantSpark, we successfully built a k-means model on the full
$2500 \times 80$ million matrix to cluster individuals by
their ethnicity achieving an Adjusted Rand Index of 0.84 (with 1 being perfect clustering  and -1 random clustering).

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  Random Forests~\cite{Breiman.2001} (RF) as an ensemble of weak learners 
\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
%http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
between features, which is of importance for modelling complex polygenic diseases.

@@

We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
excessive resource requirements on the driver or worker nodes.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the first section we

@@

As discussed in the Introduction, random forests are robust against overfitting
\marginpar{strong claim -- see disussion in Segal 2004, perhaps}
and hence can generalise for datasets where $p > n$, however here we demonstrate that this holds true in extreme cases where $p \gg n$.
